\documentclass[11pt,article,oldfontcommands]{cpp-thesis}

\setlength{\parskip}{\medskipamount}

\title{Global Value Numbering in Factor}
\author{Alex Vondrak}
\degree{Master of Science}
\field{Computer Science}
\chair{Dr. Craig Rich}{Computer Science}
\memberA{Dr. Daisy Sang}{Computer Science}
\memberB{Dr. Amar Raheja}{Computer Science}
\quarter{Summer 2011}

\input{preamble}
\input{gls/acronyms}

\begin{document}

\frontmatter

\begin{ack}
  \null
  % Bah; ack header effectively kills joke---I want a dedication, dammit
  \begin{vplace}[0.1]
    \begin{center}
      \textit{\Large To Lindsay---he is my rock}
    \end{center}
  \end{vplace}
\end{ack}

\begin{abstract}

  Compilers translate code in one programming language into semantically
  equivalent code in another language---canonically from a high-level language
  to low-level machine primitives.  Generally, the further removed a
  language's abstractions get from those of a computer, the harder it gets to
  compile code into an efficient representation.  What isn't redundant in the
  source language may map to repetitive target instructions that waste time
  recomputing results.  To combat this, compilers try to optimize away
  redundancies by looking for values that are provably equivalent when the
  program is run.

  This thesis explores the theory and implementation of a particularly
  aggressive analysis called global value numbering in a particularly
  high-level language called Factor.  Factor is a stack-based,
  dynamically-typed, object-oriented language born in late $2003$.  A baby
  among languages (now at version $0.94$), its compiler craves all the
  optimizations it can get.  By altering the existing local value numbering
  pass, redundancies can be identified and eliminated across entire programs,
  rather than isolated regions of code.  This induces speedups as high as
  $45\%$ across the majority of benchmarks.  The results from these
  comparatively simple changes hold much promise for future improvements in
  making Factor programs more efficient.
  \newpage
\end{abstract}

\tableofcontents*
\listoffigures

\mainmatter

        \input{sec/intro}
\newpage\input{sec/primer}
\newpage\input{sec/compiler}
\newpage\input{sec/vn}

\renewcommand{\bibname}{References}\newpage\printbibliography

\end{document}
