\inputlst{tree}

\begin{itemize}
\item
\begin{flushleft}
Loading constants:
\Verb|##load-integer|,
\Verb|##load-reference|
\end{flushleft}

\item
\begin{flushleft}
Optimized loading of constants, inserted by representation selection:
\Verb|##load-tagged|,
\Verb|##load-float|,
\Verb|##load-double|,
\Verb|##load-vector|
\end{flushleft}

\item
\begin{flushleft}
Stack operations:
\Verb|##peek|,
\Verb|##replace|,
\Verb|##replace-imm|,
\Verb|##inc-d|,
\Verb|##inc-r|
\end{flushleft}

\item
\begin{flushleft}
Subroutine calls:
\Verb|##call|,
\Verb|##jump|,
\Verb|##prologue|,
\Verb|##epilogue|,
\Verb|##return|
\end{flushleft}

\item
\begin{flushleft}
Inhibiting \gls{TCO}:
\Verb|##no-tco|
\end{flushleft}

\item
\begin{flushleft}
Jump tables:
\Verb|##dispatch|
\end{flushleft}

\item
\begin{flushleft}
Slot access:
\Verb|##slot|,
\Verb|##slot-imm|,
\Verb|##set-slot|,
\Verb|##set-slot-imm|
\end{flushleft}

\item
\begin{flushleft}
Register transfers:
\Verb|##copy|,
\Verb|##tagged>integer|
\end{flushleft}

\item
\begin{flushleft}
Integer arithmetic:
\Verb|##add|,
\Verb|##add-imm|,
\Verb|##sub|,
\Verb|##sub-imm|,
\Verb|##mul|,
\Verb|##mul-imm|,
\Verb|##and|,
\Verb|##and-imm|,
\Verb|##or|,
\Verb|##or-imm|,
\Verb|##xor|,
\Verb|##xor-imm|,
\Verb|##shl|,
\Verb|##shl-imm|,
\Verb|##shr|,
\Verb|##shr-imm|,
\Verb|##sar|,
\Verb|##sar-imm|,
\Verb|##min|,
\Verb|##max|,
\Verb|##not|,
\Verb|##neg|,
\Verb|##log2|,
\Verb|##bit-count|
\end{flushleft}

\item
\begin{flushleft}
Float arithmetic:
\Verb|##add-float|,
\Verb|##sub-float|,
\Verb|##mul-float|,
\Verb|##div-float|,
\Verb|##min-float|,
\Verb|##max-float|,
\Verb|##sqrt|
\end{flushleft}

\item
\begin{flushleft}
Single/double float conversion:
\Verb|##single>double-float|,
\Verb|##double>single-float|
\end{flushleft}

\item
\begin{flushleft}
Float/integer conversion:
\Verb|##float>integer|,
\Verb|##integer>float|
\end{flushleft}

\item
\begin{flushleft}
SIMD operations:
\Verb|##zero-vector|,
\Verb|##fill-vector|,
\Verb|##gather-vector-2|,
\Verb|##gather-int-vector-2|,
\Verb|##gather-vector-4|,
\Verb|##gather-int-vector-4|,
\Verb|##select-vector|,
\Verb|##shuffle-vector|,
\Verb|##shuffle-vector-halves-imm|,
\Verb|##shuffle-vector-imm|,
\Verb|##tail>head-vector|,
\Verb|##merge-vector-head|,
\Verb|##merge-vector-tail|,
\Verb|##float-pack-vector|,
\Verb|##signed-pack-vector|,
\Verb|##unsigned-pack-vector|,
\Verb|##unpack-vector-head|,
\Verb|##unpack-vector-tail|,
\Verb|##integer>float-vector|,
\Verb|##float>integer-vector|,
\Verb|##compare-vector|,
\Verb|##test-vector|,
\Verb|##test-vector-branch|,
\Verb|##add-vector|,
\Verb|##saturated-add-vector|,
\Verb|##add-sub-vector|,
\Verb|##sub-vector|,
\Verb|##saturated-sub-vector|,
\Verb|##mul-vector|,
\Verb|##mul-high-vector|,
\Verb|##mul-horizontal-add-vector|,
\Verb|##saturated-mul-vector|,
\Verb|##div-vector|,
\Verb|##min-vector|,
\Verb|##max-vector|,
\Verb|##avg-vector|,
\Verb|##dot-vector|,
\Verb|##sad-vector|,
\Verb|##horizontal-add-vector|,
\Verb|##horizontal-sub-vector|,
\Verb|##horizontal-shl-vector-imm|,
\Verb|##horizontal-shr-vector-imm|,
\Verb|##abs-vector|,
\Verb|##sqrt-vector|,
\Verb|##and-vector|,
\Verb|##andn-vector|,
\Verb|##or-vector|,
\Verb|##xor-vector|,
\Verb|##not-vector|,
\Verb|##shl-vector-imm|,
\Verb|##shr-vector-imm|,
\Verb|##shl-vector|,
\Verb|##shr-vector|
\end{flushleft}

\item
\begin{flushleft}
Scalar/vector conversion:
\Verb|##scalar>integer|,
\Verb|##integer>scalar|,
\Verb|##vector>scalar|,
\Verb|##scalar>vector|
\end{flushleft}

\item
\begin{flushleft}
Boxing and unboxing aliens:
\Verb|##box-alien|,
\Verb|##box-displaced-alien|,
\Verb|##unbox-any-c-ptr|,
\Verb|##unbox-alien|
\end{flushleft}

\item
\begin{flushleft}
Zero-extending and sign-extending integers:
\Verb|##convert-integer|
\end{flushleft}

\item
\begin{flushleft}
Raw memory access:
\Verb|##load-memory|,
\Verb|##load-memory-imm|,
\Verb|##store-memory|,
\Verb|##store-memory-imm|
\end{flushleft}

\item
\begin{flushleft}
Memory allocation:
\Verb|##allot|,
\Verb|##write-barrier|,
\Verb|##write-barrier-imm|,
\Verb|##alien-global|,
\Verb|##vm-field|,
\Verb|##set-vm-field|
\end{flushleft}

\item
\begin{flushleft}
The \gls{FFI}:
\Verb|##unbox|,
\Verb|##unbox-long-long|,
\Verb|##local-allot|,
\Verb|##box|,
\Verb|##box-long-long|,
\Verb|##alien-invoke|,
\Verb|##alien-indirect|,
\Verb|##alien-assembly|,
\Verb|##callback-inputs|,
\Verb|##callback-outputs|
\end{flushleft}

\item
\begin{flushleft}
Control flow:
\Verb|##phi|,
\Verb|##branch|
\end{flushleft}

\item
\begin{flushleft}
Tagged conditionals:
\Verb|##compare-branch|,
\Verb|##compare-imm-branch|,
\Verb|##compare|,
\Verb|##compare-imm|
\end{flushleft}

\item
\begin{flushleft}
Integer conditionals:
\Verb|##compare-integer-branch|,
\Verb|##compare-integer-imm-branch|,
\Verb|##test-branch|,
\Verb|##test-imm-branch|,
\Verb|##compare-integer|,
\Verb|##compare-integer-imm|,
\Verb|##test|,
\Verb|##test-imm|
\end{flushleft}

\item
\begin{flushleft}
Float conditionals:
\Verb|##compare-float-ordered-branch|,
\Verb|##compare-float-unordered-branch|,
\Verb|##compare-float-ordered|,
\Verb|##compare-float-unordered|
\end{flushleft}

\item
\begin{flushleft}
Overflowing arithmetic:
\Verb|##fixnum-add|,
\Verb|##fixnum-sub|,
\Verb|##fixnum-mul|,
\Verb|##save-context|
\end{flushleft}

\item
\begin{flushleft}
\Gls{GC} checks:
\Verb|##check-nursery-branch|,
\Verb|##call-gc|
\end{flushleft}

\item
\begin{flushleft}
Spills and reloads, inserted by the register allocator:
\Verb|##spill|,
\Verb|##reload|
\end{flushleft}
\end{itemize}

% dls.pdf verbatim:

The Factor implementation is structured into a virtual ma- chine (VM) written
in C++ and a core library written in Factor. The VM provides essential runtime
services, such as garbage collection, method dispatch, and a base compiler.
The rest is implemented in Factor.

The VM loads an image ﬁle containing a memory snap- shot, as in many Smalltalk
and Lisp systems. The source parser manipulates the code in the image as new
deﬁnitions are read in from source ﬁles. The source parser is written in Factor
and can be extended from user code (Section 2.3.1).  The image can be saved,
and effectively acts as a cache for compiled code.  Values are referenced using
tagged pointers [29]. Small integers are stored directly inside a pointer’s
payload. Large integers and ﬂoating point numbers are boxed in the heap;
however, compiler optimizations can in many cases elimi- nate this boxing and
store ﬂoating point temporaries in regis- ters. Specialized data structures are
also provided for storing packed binary data without boxing (Section 2.4).
Factor uses a generational garbage collection strategy to optimize workloads
which create large numbers of short- lived objects. The oldest generation is
managed using a mark-sweep-compact algorithm, with younger generations managed
by a copying collector [46]. Even compiled code is subject to compaction, in
order to reduce heap fragmentation in applications which invoke the compiler at
runtime, such as the development environment. To support early binding, the
garbage collector must modify compiled code and the callstack to point to newly
relocated code.  Run-time method dispatch is handled with polymorphic inline
caches [32]. Every dynamic call site starts out in an uninitialized cold state.
If there are up to three unique receiver types, a polymorphic inline cache is
generated for the call site. After more than three cache misses, the call site
transitions into a megamorphic call with a cache shared by all call sites.  All
source code is compiled into machine code by one of two compilers, called the
base compiler and optimizing compiler. The base compiler is a context threading
compiler implemented in C++ as part of the VM, and is mainly used for
bootstrapping purposes. The optimizing compiler is writ- ten in Factor and is
used to compile most code.  Factor is partially self-hosting and there is a
bootstrap process, similar to Steel Bank Common Lisp [38]. An im- age
generation tool is run from an existing Factor instance to produce a new
bootstrap image containing the parser, ob- ject system, and core libraries. The
Factor VM is then run with the bootstrap image, which loads a minimal set of
li- braries which get compiled with the base compiler. The op- timizing
compiler is then loaded, and the base libraries are recompiled with the
optimizing compiler. With the optimiz- ing compiler now available, additional
libraries and tools are loaded and compiled, including Factor’s GUI develop-
ment environment. Once this process completes, the image is saved, resulting in
a full development image.

The primary design considerations of the base compiler are fast compilation
speed and low implementation complexity.  As a result, the base compiler
generates context-threaded code with inlining for simple primitives [3],
performing a single pass over the input quotation.  The base compiler generates
code using a set of machine code templates for basic operations such as
creating and tearing down a stack frame, pushing a literal on the stack, making
a subroutine call, and so on. These machine code templates are generated by
Factor code during the bootstrap process. This allows the base and optimizing
compilers to share a single assembler backend written in Factor.


The optimizing compiler is structured as a series of passes operating on two
intermediate representations (IRs), referred to as high-level IR and low-level
IR. High-level IR represents control ﬂow in a similar manner to a
block-structured programming language\todo{or, y'know, like annotated stack
code}. Low-level IR represents control ﬂow with a control ﬂow graph of basic
blocks. Both intermediate forms make use of single static assignment (SSA) form
to improve the accuracy and efﬁciency of analysis [12].

High-level IR is constructed by the stack effect checker.  Macro expansion and
quotation inlining is performed by the stack checker online while high-level IR
is being con- structed. The front end does not deal with local variables, as
these have already been eliminated.

When static type information is available, Factor’s compiler can eliminate
runtime method dispatch and allocation of in- termediate objects, generating
code specialized to the under- lying data structures. This resembles previous
work in soft typing [10]. Factor provides several mechanisms to facilitate
static type propagation:

\begin{itemize}

\item Functions can be annotated as inline, causing the compiler to replace
calls to the function with the function body.

\item Functions can be hinted, causing the compiler to gener- ate multiple
specialized versions of the function, each assuming different input types, with
dispatch at the en- try point to choose the best-ﬁtting specialization for the
given inputs.  

\item Methods on generic functions propagate the type infor- mation for their
dispatched-on inputs.  

\item Functions can be declared with static input and output types using the
typed library.

\end{itemize}

The three major optimizations performed on high-level IR are sparse conditional
constant propagation (SCCP [45]), escape analysis with scalar replacement, and
overﬂow check elimination using modular arithmetic properties.  The major
features of our SCCP implementation are an extended value lattice, rewrite
rules, and ﬂow sensitivity.  Our SCCP implementation augments the standard
single- level constant lattice with information about object types, numeric
intervals, array lengths and tuple slot types. Type transfer functions are
permitted to replace nodes in the IR with inline expansions. Type functions are
deﬁned on many of the core language words.  SCCP is used to statically dispatch
generic word calls by inlining a speciﬁc method body at the call site. This
inlining generates new type information and new opportunities for constant
folding, simpliﬁcation and further inlining. In par- ticular, generic
arithmetic operations which require dynamic dispatch in the general case can be
lowered to simpler opera- tions as type information is discovered. Overﬂow
checks can be removed from integer operations using numeric interval
information. The analysis can represent ﬂow-sensitive type information.
Additionally, calls to closures which combina- tor inlining cannot eliminate
are eliminated when enough in- formation is available [16].  An escape analysis
pass is used to discover object alloca- tions which are not stored on the heap
or returned from the current function. Scalar replacement is performed on such
allocations, converting tuple slots into SSA values.  The modular arithmetic
optimization pass identiﬁes in- teger expressions in which the ﬁnal result is
taken to be modulo a power of two and removes unnecessary overﬂow checks from
any intermediate addition and multiplication operations. This novel
optimization is global and can operate over loops.

Low-level IR is built from high-level IR by analyzing control ﬂow and making
stack reads and writes explicit. During this construction phase and a
subsequent branch splitting phase, the SSA structure of high-level IR is lost.
SSA form is recon- structed using the SSA construction algorithm described in
[8], with the minor variation that we construct pruned SSA form rather than
semi-pruned SSA, by ﬁrst computing live- ness. To avoid computing iterated
dominance frontiers, we use the TDMSC algorithm from [13].  The main
optimizations performed on low-level IR are local dead store and redundant load
elimination, local value numbering, global copy propagation, representation
selec- tion, and instruction scheduling.  The local value numbering pass
eliminates common sub- expressions and folds expressions with constant operands
[9]. Following value numbering and copy propagation, a representation selection
pass decides when to unbox ﬂoating point and SIMD values. A form of instruction
scheduling intended to reduce register pressure is performed on low- level IR
as the last step before leaving SSA form [39].  We use the second-chance
binpacking variation of the lin- ear scan register allocation algorithm [43,
47]. Our variant does not take φ nodes into account, so SSA form is destruc-
ted ﬁrst by eliminating φ nodes while simultaneously per- forming copy
coalescing, using the method described in [6].
