\section{Introduction}\label{sec:intro}

% CFGs, SSA form

Compilers translate programs written in a source language (e.g., Java) into
semantically equivalent programs in some target language (e.g., assembly code).
They let us make our source language arbitrarily abstract so we can write
programs in ways that humans understand while letting the computer execute
programs in ways that machines understand.  In a perfect world, such
translation would be straightforward.  Reality, however, is unforgiving.
Straightforward compilation results in clunky target code that performs a lot
of redundant computations.  To produce efficient code, we must rely on
less-than-straightforward methods.  Typical compilers go through a stage of
\term{optimization}, whereby a number of semantics-preserving transformations
are applied to an \term{\acrlong{IR}} of the source code.  These then
(hopefully) produce a more efficient version of said representation.
Optimizers tend to work in \term{phases}, applying specific transformations
during any given phase.

\Gls{GVN} is such an analysis performed by many highly-optimizing compilers.
Its roots run deep through both the theoretical and the practical.  Using the
results of this analysis, the compiler can identify expressions in the source
code that produce the same value---not just by lexical comparison (i.e.,
variables having the same name), but by proving equivalences between what's
actually computed at runtime.  These expressions can then be simplified by
further algorithms for redundancy elimination.  This is the very essence of
most compiler optimizations: avoid redundant computation, giving us code that
runs as quickly as possible while still following what the programmer
originally wrote.

High-level, dynamic languages tend to suffer from efficiency issues: they're
often interpreted rather than compiled, and perform no heavy optimization of
the source code.  However, the Factor language (\url{http://factorcode.org})
fills an intriguing design niche, as it's very high-level yet still fully
compiled.  It's still young, though, so its compiler craves all the
improvements it can get.  In particular, while Factor currently has a
\term{local} value numbering analysis, it is inferior to \gls{GVN} in several
significant ways.

In this thesis, we explore the implementation and use of \gls{GVN} in improving
the strength of optimizations in Factor.  After establishing some preliminary
terminology and concepts in \cref{sec:intro:prelim}, we turn our attention to
the details.  Because Factor is a young and relatively unknown language,
\cref{sec:primer} provides a short tutorial, laying a foundation for
understanding the changes.  \Cref{sec:compiler} describes the overall
architecture of the Factor compiler, highlighting where the exact contributions
of this thesis fit in.  Finally, \cref{sec:vn} goes into detail about the
existing and new value numbering passes, closing with a look at the results
achieved and directions for future work.  In total, the following Factor
libraries were written for this thesis:
\begin{itemize}
  \item A \factor|graphviz| library, which provides bindings to create and
        manipulate graphs in Factor and output them using Graphviz
        (\url{http://graphviz.org}).
  \item The \factor|compiler.cfg.graphviz| library, which sues the Graphviz
        bindings to output images of Factor's low-level \acrlong{IR}.  This is
        responsible for the illustrations of optimization passes seen in
        \cref{sec:vn}.
  \item The \factor|compiler.cfg.gvn| module, which was created by copying the
        existing value numbering code (from
        \factor|compiler.cfg.value-numbering|) to make the pass global.
\end{itemize}

All code was written atop Factor version $0.94$, and copies of it can be found
in the appendices.  In the unlikely event that you want to cite this thesis,
you may use the following \textsc{Bib}\TeX~entry:
\begin{Verbatim}[gobble=2,frame=single]
  @mastersthesis{vondrak:11,
    author = {Alex Vondrak},
    title  = {Global Value Numbering in Factor},
    school = {California Polytechnic State University, Pomona},
    month  = sep,
    year   = {2011},
  }
\end{Verbatim}

\input{sec/intro/prelim}
