\chapter{Value Numbering}\label{sec:vn}

At a very basic level, most optimization techniques revolve around avoiding
redundant or unnecessary computation.  Thus, it's vital that we discover which
values in a program are equal.  That way, we can simplify the code that wastes
machine cycles repeatedly calculating the same values.  Classic optimization
phases like constant/copy propagation, common subexpression elimination,
loop-invariant code motion, induction variable elimination, and others
discussed in the de facto treatise, ``The Dragon Book'' \autocite{DragonBook},
perform this sort of redundancy elimination based on information about the
equality of expressions.

In general, the problem of determining whether two expressions in a program are
equivalent is undecidable.  Therefore, we seek a \term{conservative} solution
that doesn't necessarily identify all equivalences, but is nevertheless correct
about any equivalences it does identify.  Solving this equivalence problem is
the work of \term{value numbering} algorithms.  These assign every value in the
program a number such that two values have the same value number if and only if
the compiler can prove they will be equal at runtime.

Value numbering has a long history in literature and practice, spanning many
techniques.  In \cref{sec:compiler:cfg} we saw the \Verb|value-numbering| word,
which is actually based on some of the earliest---and least effective---methods
of value numbering.  \Cref{sec:vn:local} describes the way Factor's current
algorithm works, highlighting its shortcomings to motivate the main work of
this thesis, which is covered in \cref{sec:vn:global,sec:vn:avail}.  We finish
the \lcnamecref{sec:vn} by analyzing the results of these changes and reviewing
the literature for further enhancements that can be made to this optimization
pass.

\input{sec/vn/local}
\input{sec/vn/global}
\input{sec/vn/avail}
\input{sec/vn/results}
\input{sec/vn/future}
